---
title: "Classification and Pricing of Farmed Abalone"
author: "Amanjit Gill"
date: "2022-12-05"
output: 
    html_notebook:
        theme: flatly
---

```{css echo=FALSE}

h1, h2, h3 {
    font-weight: bold;
}

h1.title {
    font-size: 2em;
}

h2 {
    font-size: 1.5em;
}

h3 {
    font-size: 1.2em;
}

```


## 1. Exploratory Analysis

```{r setup, message=FALSE}

library(dplyr)

params <- par(
    family = "serif",
    ps = 10.5,
    mar = c(4.5, 4.5, 1, 1)
)


abalone <- read.csv("abalone.csv")

```

### 1.1 Summary statistics

Statistical summary for each numerical variable.
```{r summary}
abalone.num <- abalone %>% select(-Sex)
summary(abalone.num)

```

RESULTS:

none of the maximum values appear to be out of range, according to sources. however:

min height = 0 (impossible)

all the minimum weights are between 0.1 and 0.4 grams. not sure how realistic it is to catch and measure an abalone that tiny. needs exploration of outliers.

Check for missing values.
```{r missing values}

nrow(abalone)
sum(complete.cases(abalone))
```

RESULT: No missing values; all rows are complete.

### 1.2 Univariate outliers
```{r outliers}

pdf("1.2.pdf")
par(
    family = "serif",
    ps = 11,
    mar = c(4.5, 1, 1, 1)
)
par("mfrow"=c(4,2))

cols <- colnames(abalone.num)

for (i in 1:ncol(abalone.num)) {
    boxplot(abalone.num[,i], horizontal = TRUE, xlab = cols[i])
}
```

RESULTS:


